Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-08 paper/metamonads.tex
new intro

\usepackage[medium,compact]{titlesec}

@@

\section{Introduction}

\section{The problem of generating names, CPS, and monad}\label{CPS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% And now, a better -- and shorter -- write-up.

We will be using MetaOcaml which, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them. Figure~\ref{easycode} shows the simplest code generator |one|,
as well as more complex generators.

@@

>>>>

%\usepackage[medium,compact]{titlesec}

@@

% previous introduction is saved at the end
\section{Introduction}

In high-performance, symbolic, and numeric computing, there is a
well-known issue of balancing between maximal performance and the
level of abstraction at which code is written.  Furthermore, already
in linear algebra, there is a wealth of different aspects that
\emph{may} need to be addressed. For example, implementations of the
widely used Gaussian Elimination (GE) algorithm --- the running
example of our paper --- may need to account for the representation of
the matrix, whether to compute and return the detrminant or rank, how
and whether search for pivot, etc. Furthermore, current architectures
demand more and more frequent tweaks which, in general, cannot be done by the
compiler because the tweaking often involves domain knowledge. 
The survey \cite{carette04} of
Gaussian elimination implementation in an industrial package Maple
found XXX various aspects and XXX different implementations of the
algorithm.  We can manually write each of these implementations
optimizing for particular aspects and using cut-and-paste to ``share''
similar pieces of code. \oleg{I don't think we need to say much here
-- the shear number of various implementations should be enough of a
clue.} We can write a very generic GE procedure that accounts for
all the aspects with appropriate abstractions (cite axiom?). The
abstraction mechanisms however -- be they procedure, method or a
function call -- have a significant cost, especially for
high-performance numerical computing (cite? At least carette04).

A more appealing (acceptable? indispensable? need a better word)
approach is generative programming
\cite{Czarnecki,Veldhuizen:1998:ISCOPE,musser94algorithmoriented,BOOST,POOMA,ATLAS}.
The approach is not without problems, e.g., making sure that the
generated code is well-formed. This is a challenge in string-based
generation systems, which generally do not offer such guarantees and
therefore make it very difficult to determine which part of the
generator is at fault when the generated code cannot be parsed. Other
problems is preventing accidental variable capture (so-called hygiene
\cite{HygienicMacros}) and ensuring the generated code is
well-typed. Lisp-style macros, Scheme hygienic macros, camlp4
preprocessor (cite XXX), C++ template meta-programming, Template
Haskell \cite{conf/dagstuhl/CzarneckiOST03} solve some of the above
problems. Of the widely available maintainable languages, only
MetaOCaml \cite{CTHL03,metaocaml-org}  solves all the above problems
including the well-typing of the generated code \cite{TahaSheard97,TahaThesis}.

But more difficult problems remain. Is the generated code optimal? Do
we still need post-processing to eliminate common subexpressions and
fold constants, remove redundant bindings? Is the generator readable,
resembling the original algorithm, and extensible? Are the aspects
truly modular? Can we add another aspect to it or another instance of
the existing aspect without affecting the existing ones? Finally, can
we express domain-specific knowledge, e.g., one should not attempt to
use full division when dealing with matrices of exact integers, nor is
it worthwhile to use full pivoting on a matrix of rational integers
($\mathbb Q $).

MetaOCaml is \emph{generative}: generated code can only be treated as
a black box: it cannot be inspected and it cannot be post-processed
(i.e., no intensional analysis). This approach gives a stronger
equational theory \cite{Taha2000}, and avoids the danger of creating
unsoundness \cite{TahaThesis}. Furthermore, intensional code analysis
essentially requires one to insert both an optimizing compiler and an
automated theorem proving system into the code generating system
\cite{Pueschel:05,Kennedy01Telescoping,dongarra7,Veldhuizen:2004}.
While this is potentially extremely powerful and an exciting area of
research, it is also extremely complex, which means that it is
currently more error-prone and difficult to ascertain the correctness
of the resulting code.

Therefore, in MetaOCaml, code must be generated just right (see
\cite{TahaThesis} for many simple examples).  For more complex
examples, new techniques are necessary, e.g., abstract interpretation
\cite{KiselyovTaha}.  But more problems remain
\cite{Padua:MetaOcaml:04}: generating binding statements (``names''),
especially when generating loop bodies or conditional branches; making
continuation-passing style (CPS) code clear.  Many authors
understandably shy away from CPS code as it quickly becomes
unreadable.  But this is needed for proper name generation.
The problems of compositionality of code generators, expressing
dependencies among them and domain-specific knowledge remain.

In this paper we report on progress of solving these problems using GE
as our running example. Specifically, our contributions:
\vspace*{-6pt}
\begin{itemize}
    \item Extending monad of \cite{KiselyovTaha} for generating
      binding statements when generating control structures such as
      loops and conditionals. We argue that code generation is an
      effect and so requires something like a monad for proper
      serialization.
    \item Implementation of the |mdo|-notation (patterned after
      |do|-notation of Haskell) to make monadic code readable
    \item Use of functors (including higher-order functors) to
      modularize the generator, express aspects (including results of
      various types) and \emph{assure composability of aspects} even
      for aspects that use state and have to be accounted in many
      places in the generated code
    \item use functor type sharing constraints to encode domain-specific
      knowledge
\end{itemize}

The rest of this paper is structured as follows: The next section
introduces code generation in MetaOCaml, the problem of name
generation, and continuation-passing style (CPS) as a general
solution.  We also introduce the monad and the issues of generating
control statements. Section~\ref{functors} describes the use of
parametrized modules of OCaml to encode all of the aspects of the
Gaussian Elimination algorithm family in completely separate,
independent modules.  We briefly discuss related work in
section~\ref{related}. We then outline the future work and conclude.
Appendices give samples of the generated code (which is availablke in
full at XXX).

We wish to thank Cristiano Calgano for his help in adapting camlp4 for
use with MetaOCaml.

\section{Generating binding statements, CPS, and monad}\label{CPS}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

We build code generators out of primitive ones using code generation 
combinators. MetaOcaml, as an instance of a multi-stage
programming system \cite{TahaThesis}, provides exactly the needed
features: to construct a code expression, to combine them, and to
execute them. Figure~\ref{easycode} shows the simplest code generator |one|,
as well as more complex generators.

@@

% old intro
\section{Introduction}
